High-accuracy neutron diffusion calculations based on integral transport theory

### 2019-12-09 docs/EPJP2019/Gross-EPJP-2019.tex
partial advancement of revision by Tomatis

The starting point for the derivation of the integral expression for the flux in slab geometry is to directly integrate along a line the transport equation~\cite{Davison-1957,Pomraning-1973,Duderstadt-1979,Lewis-1984}
\begin{equation}\label{eq:full-ITE-w-bndry-terms}
\psi(\rr,\OO) = \int_0^R { dR'
	Q(\rr-R'\OO,\OO) e^{-\mcL(\rr,\rr-R'\OO)}
} + \psi(\rr-R\OO,\OO) e^{-\mcL(\rr,\rr-R\OO)},
\end{equation}
where $\mcL(\rr,\rr-R'\OO)$ is the \emph{optical length}, defined as
\begin{equation}\label{eq:opl}
\mcL(\rr,\rr-R'\OO) \equiv \int_0^{R'} {\sigma(\rr-R''\OO)dR'} .
\end{equation}
Assuming isotropic scattering and homogeneous medium in one-dimension for~\eq{eq:full-ITE-w-bndry-terms} recovers~\eq{eq:flux-slab-l-0-no-bndry-terms}.

@@

Only scattering in isotropic media is considered here: $\sigma_s(\ast, \OO' \rightarrow \OO) = \sigma_s(\ast, \mu_0)/2\pi$ with the angle cosine $\mu_0 = \OO' \cdot \OO$. The source terms are written by moments of Legendre polynomials after the usual expansion of the scattering cross sections in Legendre Polynomials of $\mu_0$, followed by the application of the addition theoreom and by integration on spherical harmonics using the symmetries and rotational invariance of the slab geometry~\cite{Bell-1970}. The cross sections of the source term are grouped for simplicity by introducing the $l$-th moments of the production cross section~\cite{Tomatis-2011}
\begin{equation}\label{eq:src-mmnts}
\sigma_l(x,E\leftarrow E') = \sigma_{s,l}(x,E\leftarrow E')
+ \delta_{l0}\frac{\chi(E)}{\keff}\nu\sigma_f(x,E'),
\end{equation}
with the Legendre moments of the scattering cross section
\begin{equation}\label{eq:src-mmnts-explicit}
\sigma_{s,l}(x,E\leftarrow E') = \int_{-1}^1 d\mu_0 P_l(\mu_0)\sigma_s(x,E\leftarrow E',\mu_0).
\end{equation}
$\delta_{l0}$ in~\eq{eq:src-mmnts} is the Kronecker function equal to 1 only for $l=0$ and zero otherwise. The multiplication factor $\keff$ applies here on the fission production since external sources are missing. The source term is then rewritten as
% Daniele: the following equation could be simplified with fewer lines thank to the previous eqs.
\begin{equation}\label{eq:src}
\begin{split}
Q(x,E,\mu)=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_{s,l}(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} 
+ \frac{\chi(E)}{2\keff} \int_0^\infty { dE'
	\nu\sigma_f(x,E') \psi_0(x,E')
} \\
=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_l(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} \\
=&\sum\limits_{l=0}^\infty {
\frac{2l+1}{2} \int_0^\infty { dE'
	q_l(x,E\leftarrow E')\pl{\mu} 			
	}
} \\
=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} q_l(x,E)\pl{\mu} 			
},
\end{split}
\end{equation}
where the source angular moments are defined as
\begin{equation}\label{eq:q_l}
q_l(x,E) \equiv \int_0^\infty dE' q_l(x,E\leftarrow E') \equiv 
\int_0^\infty dE' \sigma_l(x,E\leftarrow E')\psi_l(x,E') dE' .
\end{equation}

Finally, we also expand the angular dependence of the uncollided boundary flux in a series of orthogonal Legendre polynomials:
%\begin{eqnarray}\label{eq:Q-series}
%\psi(0,E,\mu) &=& \sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} \psi_l(0,E) \pl{\mu} 	
%} \nonumber \\
%\psi(a,E,\mu) &=& \sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} \psi_l(a,E) \pl{\mu} 	
%}.
%\end{eqnarray}
\begin{equation}\label{eq:Q-series}
\psi(x,E,\mu) = \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \pl{\mu} \psi_l(x,E)
}.
\end{equation}
This allows dealing only with flux moments at the right hand side of~\eqs{eq:src-mmnts}.

@@

%The independent energy parameter is omitted here for brevity and clarity but restoring it is straightforward.
%
The scalar flux $\phi^+$ at $x$ resulting from neutrons coming from the left, i.e., $x'<x$ $(\mu>0)$, is determined by substituting Eqs.~\eqref{eq:Q-series} in Eq.~\eqref{eq:int-ang-flx-a} and integrating over the angle 
\begin{equation}\label{eq:phi+}
\begin{split}
\phi^+(x,E) \equiv& \int_0^1 {d\mu \psi(x,E,\mu)} \\
%&=& \int_0^1 {d\mu \psi(0,\mu) e^{-\mcL(0,x)/\mu} } 
%+ \int_0^1 {d\mu 
%	\int_0^x { dx'
%		\frac{Q(x',\mu)}{\mu} e^{-\mcL(x',x)/\mu} 
%	}
%}
%\nonumber \\
=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} 
		\left\{\psi_l(0,E) \int_0^1 { d\mu
		\pl{\mu} e^{-\mcL(0,x,E)/\mu}	
	}
	+	\int_0^x { dx'
		q_l(x',E)
		\int_0^1 { d\mu
			\frac{\pl{\mu}}{\mu} e^{-\mcL(x',x,E)/\mu}	
		}
	}
\right\}
}.
\end{split}
\end{equation}
In order to solve the angular integrals we resort to the following result~\cite{Cengel-1984,Settle-1994}
\begin{equation}\label{eq:Sni}
\int_0^1 { 
	\mu^i \pl{\mu} e^{-y/\mu} d\mu
} = \sum\limits_{m=0}^{[l/2]} {
	h_m E_{l+2+i-2m}(y)
},
\end{equation}
where the coefficients $h_m$ are used to rewrite the Legendre polynomials as simple sums of weighted monomials, like
\begin{equation}\label{eq:h_i}
\pl{x}=\sum\limits_{m=0}^{[l/2]} {
	h_m x^{n-2m}
}, \quad
% \end{equation}
% and given explicitly by
% \begin{equation}\label{eq:h_i-explicit}
h_m = \frac{(-1)^m (2l-2m)!}{2^l m! (l-m)! (l-2m)!}
%\end{equation}
%where
%\begin{equation}\label{eq:l_2}
\quad \text{and} \quad
[l/2] = 
\begin{cases}
l/2, & l \text{ even}, \\
(l - 1)/2, & l \text{ odd}.
\end{cases}
\end{equation}
%
Hence, the contribution to the flux from $x'<x$ can be written as
\begin{equation}\label{eq:phi+2}
\phi^+(x)= \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \sum\limits_{m=0}^{[l/2]} { h_m
		\left\{                
		        E_{l+2-2m}[\mcL(0,x,E)]                         
		        \psi_l(0,E) 
		       +       \int_0^x { dx'
		        E_{l+1-2m}[\mcL(x',x,E)]
		        q_l(x',E)                               
		} 
	\right\}
	}
}.
\end{equation}
The contribution to the flux from $x'>x$ ($\mu<0$) can be calculated in a similar manner, using the change of variables $\eta=-\mu$ for the angular integrals and the fact that $\pl{-\mu}=(-1)^l\pl{\mu}$. Hence, the expression for the scalar flux which accounts for anisotropic scattering and heterogeneous medium is
\begin{equation}\label{eq:int-scalar-flux}
\begin{split}
\phi(x,E) =& \phi^+(x,E)+\phi^-(x,E) \\
=& \sum\limits_{l=0}^\infty  \frac{2l+1}{2} 
\sum\limits_{m=0}^{[l/2]} h_m
\bigg\{
E_{l+2-2m}[\mcL(0,x,E)]	\psi_l(0,E) 
+(-1)^l E_{l+2-2m}[\mcL(x,a,E)] \psi_l(a,E) \\
&+ \int_0^x dx' E_{l+1-2m}[\mcL(x',x,E)] q_l(x',E)
      +(-1)^l \int_x^a dx' 
      E_{l+1-2m}[\mcL(x,x',E)] q_l(x',E)
\bigg\}.
\end{split}
\end{equation}

%Defining $q_l(x,E)\equiv \int_0^\infty \!\!\!\! dE' \sigma_l(x,E\leftarrow E') \psi_l(x,E')$ 
%
The generalization of~\eq{eq:int-scalar-flux} to current is obtained in a similar procedure according to
\begin{equation}\label{eq:tot-J}
\begin{split}
J(x,E) =& J^+(x,E) - J^-(x,E) = \int_0^1 {d\mu \mu \psi(x,E,\mu)} 
- \int_0^{-1} {d\mu \mu \psi(x,E,\mu)} \\
=&\sum\limits_{l=0}^\infty  \frac{2l+1}{2}
\sum\limits_{m=0}^{[l/2]}  h_m
\bigg\{
E_{l+3-2m}[\mcL(0,x,E)]	\psi_l(0,E) 
+(-1)^{l+1} E_{l+3-2m}[\mcL(x,a,E)]\psi_l(a,E) \\
&+
\int_0^x dx' E_{l+2-2m}[\mcL(x',x,E)] q_l(x',E)
+(-1)^{l+1} \int_x^a dx' E_{l+2-2m}[\mcL(x,x',E)] q_l(x',E)
\bigg\}.
\end{split}
\end{equation}
%
%For isotropic scattering ($l=0$), \eqsand{eq:int-scalar-flux}{eq:tot-J} reduce to the known expressions derived from the Peierls equation
%\begin{eqnarray}\label{eq:iso-flx-J}
%\ftr(x,E) &=& \frac{1}{2}\left\{\psi_0^0(E) E_2[\mcL(0,x,E)] 
%+ \psi_0^a E_2[\mcL(x,a,E)]\right\}
%+ \frac{1}{2}\int_0^a { dx' q_0(x',E)
%	 E_1\left[\lvert\mcL(x',x,E)\rvert\right]} , \nonumber \\
%\jtr(x,E) &=& \frac{1}{2}\left\{\psi_0^0(E) E_3[\mcL(0,x,E)] 
%- \psi_0^a E_3[\mcL(x,a,E)]\right\}
%+ \frac{1}{2}\int_0^a { dx' q_0(x',E)
%	 E_2\left[\lvert\mcL(x',x,E)\rvert\right]
%	sign(x-x') } .
%\end{eqnarray} 
%

>>>>

Another starting point for the derivation of the integral expression for the flux in slab geometry is to directly integrate along a line the transport equation~\cite{Davison-1957,Pomraning-1973,Duderstadt-1979,Lewis-1984}
\begin{equation}\label{eq:full-ITE-w-bndry-terms}
\psi(\rr,\OO) = \int_0^R { dR'
	Q(\rr-R'\OO,\OO) e^{-\mcL(\rr,\rr-R'\OO)}
} + \psi(\rr-R\OO,\OO) e^{-\mcL(\rr,\rr-R\OO)},
\end{equation}
where $\mcL(\rr,\rr-R'\OO)$ is the \emph{optical length}, defined as
\begin{equation}\label{eq:opl}
\mcL(\rr,\rr-R'\OO) \equiv \int_0^{R'} {\sigma(\rr-R''\OO)dR'} .
\end{equation}
Assuming isotropic scattering and homogeneous medium in one-dimension for~\eq{eq:full-ITE-w-bndry-terms} recovers~\eq{eq:flux-slab-l-0-no-bndry-terms}.

@@

The angular dependence of the source terms and of the uncollided boundary flux is expanded in a series of orthogonal Legendre polynomials as a consequence of the expansion of the scattering term in a series of orthogonal spherical harmonics, simply reducing to Legendre polynomials in slab geometry. For instance, for the angular flux it is:
%\begin{eqnarray}\label{eq:Q-series}
%\psi(0,E,\mu) &=& \sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} \psi_l(0,E) \pl{\mu} 	
%} \nonumber \\
%\psi(a,E,\mu) &=& \sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} \psi_l(a,E) \pl{\mu} 	
%}.
%\end{eqnarray}
\begin{equation}\label{eq:Q-series}
\psi(x,E,\mu) = \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \pl{\mu} \psi_l(x,E)
}.
\end{equation}
The cross sections in the source terms are written explicitly as~\cite{Tomatis-2011}
\begin{equation}\label{eq:src-mmnts}
\sigma_l(x,E\leftarrow E') = \sigma_{s,l}(x,E\leftarrow E')
+ \delta_{l0}\frac{\chi(E)}{\keff}\nu\sigma_f(x,E'),
\end{equation}
with the $l$-th Legendre moments of the scattering cross section
\begin{equation}\label{eq:src-mmnts-explicit}
\sigma_{s,l}(x,E\leftarrow E') = \int_{-1}^1 d\mu' P_l(\mu')\sigma_s(x,E\leftarrow E',\mu_0),
\end{equation}
where $\mu_0=\OO'\cdot\OO$ is the cosine between the incoming and outgoing directions $\OO'$ and $\OO$ in a collision. The source term is then expanded as
\begin{eqnarray}\label{eq:src}
Q(x,E,\mu)&=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_{s,l}(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} 
+ \frac{\chi(E)}{2\keff} \int_0^\infty { dE'
	\nu\sigma_f(x,E') \psi_0(x,E')
}  \nonumber \\
&=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_l(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} \nonumber \\
&=&\sum\limits_{l=0}^\infty {
\frac{2l+1}{2} \int_0^\infty { dE'
	q_l(x,E\leftarrow E')\pl{\mu} 			
	}
} \nonumber \\
&=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} q_l(x,E)\pl{\mu} 			
} 
\,\, ,
\end{eqnarray}
where the source angular moments are defined as
\begin{equation}\label{eq:q_l}
q_l(x,E) \equiv \int_0^\infty dE' q_l(x,E\leftarrow E') \equiv 
\int_0^\infty dE' \sigma_l(x,E\leftarrow E')\psi_l(x,E') dE' .
\end{equation}

@@

%The independent energy parameter is omitted here for brevity and clarity but restoring it is straightforward.
%
In order to calculate the scalar flux $\phi^+$ at $x$ resulting from neutrons coming from the left, i.e., $x'<x$ ($\mu>0$), substitute Eqs.~\eqref{eq:Q-series} in Eq.~\eqref{eq:int-ang-flx-a} and integrate over the angle 
\begin{eqnarray}\label{eq:phi+}
\phi^+(x,E)&\equiv& \int_0^1 {d\mu \psi(x,E,\mu)} \nonumber \\
%&=& \int_0^1 {d\mu \psi(0,\mu) e^{-\mcL(0,x)/\mu} } 
%+ \int_0^1 {d\mu 
%	\int_0^x { dx'
%		\frac{Q(x',\mu)}{\mu} e^{-\mcL(x',x)/\mu} 
%	}
%}
%\nonumber \\
&=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} 
		\left\{\psi_l(0,E) \int_0^1 { d\mu
		\pl{\mu} e^{-\mcL(0,x,E)/\mu}	
	}
	+	\int_0^x {\!\!\! dx'
		q_l(x',E)
		\int_0^1 { d\mu
			\frac{\pl{\mu}}{\mu} e^{-\mcL(x',x,E)/\mu}	
		}
	}
\right\}
} .
\end{eqnarray}

In order to solve the angular integrals we resort to the following result~\cite{Cengel-1984,Settle-1994}
\begin{equation}\label{eq:Sni}
\int_0^1 { 
	\mu^i \pl{\mu} e^{-y/\mu} d\mu
}
= \sum\limits_{m=0}^{[l/2]} {
	h_m E_{l+2+i-2m}(y)
} ,
\end{equation}
where $h_m$ is defined according to 
\begin{equation}\label{eq:h_i}
\pl{x}=\sum\limits_{m=0}^{[l/2]} {
	h_m x^{n-2m}
}
\end{equation}
and given explicitly by
\begin{equation}\label{eq:h_i-explicit}
h_m = \frac{(-1)^m (2l-2m)!}{2^l m! (l-m)! (l-2m)!}
\end{equation}
where
\begin{equation}\label{eq:l_2}
[l/2] = 
\begin{cases}
l/2, & l \text{ even}, \\
(l - 1)/2, & l \text{ odd}.
\end{cases}
\end{equation}
%
Hence, the contribution to the flux from $x'<x$ can be written as
\begin{equation}\label{eq:phi+2}
\phi^+(x)= \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \sum\limits_{m=0}^{[l/2]} { h_m
		\left\{                
		        E_{l+2-2m}[\mcL(0,x,E)]                         
		        \psi_l(0,E) 
		       +       \int_0^x { dx'
		        E_{l+1-2m}[\mcL(x',x,E)]
		        q_l(x',E)                               
		} 
	\right\}
	}
}.
\end{equation}
The contribution to the flux from $x'>x$ ($\mu<0$) can be calculated in a similar manner, using the change of variables $\eta=-\mu$ for the angular integrals and the fact that $\pl{-\mu}=(-1)^l\pl{\mu}$, resulting in an expression for the scalar flux which accounts for anisotropic scattering and heterogeneous medium
\begin{eqnarray}\label{eq:int-scalar-flux}
\phi(x,E) &=& \phi^+(x,E)+\phi^-(x,E) \nonumber \\
&=& 
\sum\limits_{l=0}^\infty  \frac{2l+1}{2} 
\sum\limits_{m=0}^{[l/2]}  h_m
\bigg\{
E_{l+2-2m}[\mcL(0,x,E)]	\psi_l(0,E) 
+(-1)^l E_{l+2-2m}[\mcL(x,a,E)] \psi_l(a,E) 
\nonumber \\
&&+
		\int_0^x dx' E_{l+1-2m}[\mcL(x',x,E)] q_l(x',E)
		+(-1)^l \int_x^a dx' 
		E_{l+1-2m}[\mcL(x,x',E)] q_l(x',E)
\bigg\}	\,\, .		
\end{eqnarray}

%Defining $q_l(x,E)\equiv \int_0^\infty \!\!\!\! dE' \sigma_l(x,E\leftarrow E') \psi_l(x,E')$ 
%
The generalization of~\eq{eq:int-scalar-flux} to current is obtained in a similar procedure according to
\begin{eqnarray}\label{eq:tot-J}
J(x,E) &=& J^+(x,E) - J^-(x,E) = \int_0^1 {d\mu \mu \psi(x,E,\mu)} 
- \int_0^{-1} {d\mu \mu \psi(x,E,\mu)} \nonumber \\
&=&\sum\limits_{l=0}^\infty  \frac{2l+1}{2}
\sum\limits_{m=0}^{[l/2]}  h_m
\bigg\{
E_{l+3-2m}[\mcL(0,x,E)]	\psi_l(0,E) 
+(-1)^{l+1} E_{l+3-2m}[\mcL(x,a,E)]\psi_l(a,E)  \nonumber \\
&&+
\int_0^x dx' E_{l+2-2m}[\mcL(x',x,E)] q_l(x',E)
+(-1)^{l+1} \int_x^a dx' E_{l+2-2m}[\mcL(x,x',E)] q_l(x',E)
\bigg\}	\,\, . 
\end{eqnarray}
%
%For isotropic scattering ($l=0$), \eqsand{eq:int-scalar-flux}{eq:tot-J} reduce to the known expressions derived from the Peierls equation
%\begin{eqnarray}\label{eq:iso-flx-J}
%\ftr(x,E) &=& \frac{1}{2}\left\{\psi_0^0(E) E_2[\mcL(0,x,E)] 
%+ \psi_0^a E_2[\mcL(x,a,E)]\right\}
%+ \frac{1}{2}\int_0^a { dx' q_0(x',E)
%	 E_1\left[\lvert\mcL(x',x,E)\rvert\right]} , \nonumber \\
%\jtr(x,E) &=& \frac{1}{2}\left\{\psi_0^0(E) E_3[\mcL(0,x,E)] 
%- \psi_0^a E_3[\mcL(x,a,E)]\right\}
%+ \frac{1}{2}\int_0^a { dx' q_0(x',E)
%	 E_2\left[\lvert\mcL(x',x,E)\rvert\right]
%	sign(x-x') } .
%\end{eqnarray} 
%
